---
title: "diig-fin"
output: html_document
date: "2025-01-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DIIG Data Challenge 2025: Life Expectancy

### Liane Ma

## Data

```{r load data}
library(tidyverse)
library(dplyr)
library(tibble)

life_exp <- read_csv("data/Life Expectancy Data.csv")

#cleaning
colnames(life_exp) <- gsub(" ", "_", colnames(life_exp))
colnames(life_exp) <- gsub("-", "_", colnames(life_exp))
colnames(life_exp) <- gsub("__", "_", colnames(life_exp))

life_exp_clean <- na.omit(life_exp)
```

## Exploratory Data Analysis

```{r}
life_exp_clean |>
  ggplot(aes(x = Life_expectancy))+
  geom_histogram(binwidth = 2) + 
  labs(
    title = "Distribution of Life Expectancy",
    y = "Count",
    x = "Life Expectancy (years)"
  )
```

```{r}
life_exp_clean |>
  ggplot(aes(x = Status, y = Life_expectancy)) + 
  geom_boxplot()

```

```{r}
life_exp_clean |>
  ggplot(aes(x = Alcohol, y = Life_expectancy)) + 
  geom_point(alpha = 0.5) +
  facet_wrap(~Status)
```

```{r}
life_exp_clean |>
  ggplot(aes(x = percentage_expenditure, y = Life_expectancy)) +
  geom_point(alpha = 0.5) +
  geom_smooth() + 
  facet_wrap(~Status)
```

## Methodology 

Correlation Analysis:

```{r}
correlations <- cor(life_exp_clean[, c("Adult_Mortality", "infant_deaths", "Alcohol", 
                                       "percentage_expenditure", "Hepatitis_B", "Measles", 
                                       "BMI", "under_five_deaths", "Polio", "Total_expenditure", 
                                       "Diphtheria", "HIV/AIDS", "GDP", "Population", 
                                       "thinness_1_19_years", "thinness_5_9_years", 
                                       "Income_composition_of_resources", "Schooling")], 
                    life_exp_clean$Life_expectancy)

correlations_named <- setNames(correlations, 
                               c("Adult_Mortality", "infant_deaths", "Alcohol", 
                                 "percentage_expenditure", "Hepatitis_B", "Measles", 
                                 "BMI", "under_five_deaths", "Polio", "Total_expenditure", 
                                 "Diphtheria", "HIV/AIDS", "GDP", "Population", 
                                 "thinness_1_19_years", "thinness_5_9_years", 
                                 "Income_composition_of_resources", "Schooling"))

correlations_tbl <- enframe(correlations_named, name = "Variable", value = "Correlation") %>%
  arrange(desc(Correlation))  # Sort by descending correlation

positive_corr <- correlations_tbl %>% filter(Correlation > 0)
negative_corr <- correlations_tbl %>% filter(Correlation < 0)
weak_corr <- correlations_tbl %>% filter(abs(Correlation) < 0.1)

cat("Strongest Positive Correlations with Life Expectancy:\n")
positive_corr

cat("\nStrongest Negative Correlations with Life Expectancy:\n")
negative_corr

cat("\nWeak Correlations (Close to 0):\n")
weak_corr

```

by Status:

```{r}
developed <- life_exp_clean %>% filter(Status == "Developed")
developing <- life_exp_clean %>% filter(Status == "Developing")

correlations_dev <- cor(developed[, c("Adult_Mortality", "infant_deaths", "Alcohol", 
                                      "percentage_expenditure", "Hepatitis_B", "Measles", 
                                      "BMI", "under_five_deaths", "Polio", "Total_expenditure",                                       "Diphtheria", "HIV/AIDS", "GDP", "Population", 
                                      "thinness_1_19_years", "thinness_5_9_years", 
                                      "Income_composition_of_resources", "Schooling")], 
                       developed$Life_expectancy)

correlations_devp <- cor(developing[, c("Adult_Mortality", "infant_deaths", "Alcohol", 
                                        "percentage_expenditure", "Hepatitis_B", "Measles", 
                                        "BMI", "under_five_deaths", "Polio", "Total_expenditure", 
                                        "Diphtheria", "HIV/AIDS", "GDP", "Population", 
                                        "thinness_1_19_years", "thinness_5_9_years", 
                                        "Income_composition_of_resources", "Schooling")], 
                         developing$Life_expectancy)

correlations_named_dev <- setNames(correlations_dev, 
                                   c("Adult_Mortality", "infant_deaths", "Alcohol", 
                                     "percentage_expenditure", "Hepatitis_B", "Measles", 
                                     "BMI", "under_five_deaths", "Polio", "Total_expenditure", 
                                     "Diphtheria", "HIV/AIDS", "GDP", "Population", 
                                     "thinness_1_19_years", "thinness_5_9_years", 
                                     "Income_composition_of_resources", "Schooling"))

correlations_named_devp <- setNames(correlations_devp, 
                                    c("Adult_Mortality", "infant_deaths", "Alcohol", 
                                      "percentage_expenditure", "Hepatitis_B", "Measles", 
                                      "BMI", "under_five_deaths", "Polio", "Total_expenditure", 
                                      "Diphtheria", "HIV/AIDS", "GDP", "Population", 
                                      "thinness_1_19_years", "thinness_5_9_years", 
                                      "Income_composition_of_resources", "Schooling"))

correlations_tbl_dev <- enframe(correlations_named_dev, name = "Variable", value = "Correlation_Developed") %>%
  arrange(desc(Correlation_Developed))

correlations_tbl_devp <- enframe(correlations_named_devp, name = "Variable", value = "Correlation_Developing") %>%
  arrange(desc(Correlation_Developing))
correlations_tbl_dev
correlations_tbl_devp


correlation_comparison <- left_join(correlations_tbl_dev, correlations_tbl_devp, by = "Variable")



cat("\n--- Comparison of Correlations with Life Expectancy: Developed vs Developing Countries ---\n")
print(correlation_comparison)

```

For developed countries, income, percentage expentiure, and GDP showed positive, strong correlation with life expectancy. polio, bmi, diphtheria and under-five-deaths all had a weak correlation with life expectancy with pearson coefficients \~0.

in contrast, developing countries had a strong, positive correlation between life expectancy and schooling, income comp, BMI, GDP, and percentage-expenditure, showing that for developing countries, BMI is still an important factor that plays a role in life expectancy, but that this correlation strength decreases as a country becomes a developed country. interestingly, total-expenditure had a week correlation in developing countries, and various diseases and alcohol had a mildly strong correlation of around 0.2.

therefore:

developing

-   recommend focusing on **education, BMI**

-   **secondary areas of focus include diseases such as diphtheria, polio, alcohol.**

developed

-   less emphasis on diseases such as Polio, BMI, Diphtheria etc.

both

-   income comp, percentage_expenditure, GDP, schooling,

Multiple Linear Reg Models

```{r}
# Model for developed countries
model_developed <- lm(Life_expectancy ~ GDP + Schooling + BMI + Alcohol + Adult_Mortality + 
                        infant_deaths + Polio + Total_expenditure + Diphtheria + Income_composition_of_resources, 
                      data = developed)

# Model for developing countries
model_developing <- lm(Life_expectancy ~ GDP + Schooling + BMI + Alcohol + Adult_Mortality + 
                         infant_deaths + Polio + Total_expenditure + Diphtheria + 
                         Income_composition_of_resources, 
                       data = developing)
# Summary of the model for developed countries
summary(model_developed)

# Summary of the model for developing countries
summary(model_developing)
```

developed;

-   GDP, BMI, Alcohol, infant deaths, Polio, total exp, diphtheria is not significant

-   Schooling, Adult mortality, diphtheria, income comp significant

developing: GDP, Schooling, BMI, Alcohol, Adult mortality, diphtheria, income comp are all signficant
