---
title: "diig-fin"
output: html_document
date: "2025-01-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DIIG Data Challenge 2025: Life Expectancy Dataset

### Liane Ma

## Data

```{r load data}
library(tidyverse)
library(dplyr)
library(tibble)

life_exp <- read_csv("data/Life Expectancy Data.csv")

#cleaning
colnames(life_exp) <- gsub(" ", "_", colnames(life_exp))
colnames(life_exp) <- gsub("-", "_", colnames(life_exp))
colnames(life_exp) <- gsub("__", "_", colnames(life_exp))

life_exp_clean <- na.omit(life_exp)
```

## Exploratory Data Analysis

```{r}
life_exp_clean |>
  ggplot(aes(x = Life_expectancy))+
  geom_histogram(binwidth = 2) + 
  labs(
    title = "Distribution of Life Expectancy",
    y = "Count",
    x = "Life Expectancy (years)"
  ) + 
  theme_minimal()
```

```{r}
life_exp_clean |>
  ggplot(aes(x = Status, y = Life_expectancy)) + 
  geom_boxplot() + 
  labs(
    title = "Distribution of life expectancy by developed vs. developing country",
    y = "life expectancy"
  ) + 
  theme_minimal()

```

```{r}
life_exp_clean |>
  ggplot(aes(x = Alcohol, y = Life_expectancy)) + 
  geom_point(alpha = 0.5) +
  facet_wrap(~Status) + 
  labs(
    title = "Life Expectancy vs. Alcohol per capita by Country Status", 
    y = "Life Expectancy",
    x = "Alcohol per capita consumption (litres)"
  ) + 
  theme_minimal()
```

```{r}
life_exp_clean |>
  ggplot(aes(x = percentage_expenditure, y = Life_expectancy)) +
  geom_point(alpha = 0.5) +
  geom_smooth() + 
  facet_wrap(~Status) + 
  labs(
    title = "Life Expectancy vs. Percent of Expenditure on Health", 
    subtitle = "by country status",
    y = "Life expectancy",
    x = "%GDP of Expenditure on Health"
  ) +
  theme_minimal()
```

```{r}
life_exp_clean |>
  ggplot(aes(x = Schooling, y = Life_expectancy, color = Status)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Life Expectancy vs. Schooling", 
    subtitle = "by country status",
    y = "Life expectancy",
    x = "Schooling (years)"
  ) +
  theme_minimal()
```

## Methodology

#### Correlation Analysis (general)

```{r}
correlations <- cor(life_exp_clean[, c("Adult_Mortality", "infant_deaths", "Alcohol", 
                                       "percentage_expenditure", "Hepatitis_B", "Measles", 
                                       "BMI", "under_five_deaths", "Polio", "Total_expenditure", 
                                       "Diphtheria", "HIV/AIDS", "GDP", "Population", 
                                       "thinness_1_19_years", "thinness_5_9_years", 
                                       "Income_composition_of_resources", "Schooling")], 
                    life_exp_clean$Life_expectancy)

correlations_named <- setNames(correlations, 
                               c("Adult_Mortality", "infant_deaths", "Alcohol", 
                                 "percentage_expenditure", "Hepatitis_B", "Measles", 
                                 "BMI", "under_five_deaths", "Polio", "Total_expenditure", 
                                 "Diphtheria", "HIV/AIDS", "GDP", "Population", 
                                 "thinness_1_19_years", "thinness_5_9_years", 
                                 "Income_composition_of_resources", "Schooling"))

correlations_tbl <- enframe(correlations_named, name = "Variable", value = "Correlation") %>%
  arrange(desc(Correlation))

positive_corr <- correlations_tbl |> filter(Correlation > 0)
negative_corr <- correlations_tbl |> filter(Correlation < 0)
weak_corr <- correlations_tbl |> filter(abs(Correlation) < 0.1)

cat("Strongest Positive Correlations with Life Expectancy:\n")
positive_corr

cat("\nStrongest Negative Correlations with Life Expectancy:\n")
negative_corr

cat("\nWeak Correlations (Close to 0):\n")
weak_corr

```

#### Correlation Analysis by country status

```{r}
developed <- life_exp_clean |> filter(Status == "Developed")
developing <- life_exp_clean |> filter(Status == "Developing")

correlations_dev <- cor(developed[, c("Adult_Mortality", "infant_deaths", "Alcohol", 
                                      "percentage_expenditure", "Hepatitis_B", "Measles", 
                                      "BMI", "under_five_deaths", "Polio", "Total_expenditure",                                       "Diphtheria", "HIV/AIDS", "GDP", "Population", 
                                      "thinness_1_19_years", "thinness_5_9_years", 
                                      "Income_composition_of_resources", "Schooling")], 
                       developed$Life_expectancy)

correlations_devp <- cor(developing[, c("Adult_Mortality", "infant_deaths", "Alcohol", 
                                        "percentage_expenditure", "Hepatitis_B", "Measles", 
                                        "BMI", "under_five_deaths", "Polio", "Total_expenditure", 
                                        "Diphtheria", "HIV/AIDS", "GDP", "Population", 
                                        "thinness_1_19_years", "thinness_5_9_years", 
                                        "Income_composition_of_resources", "Schooling")], 
                         developing$Life_expectancy)

correlations_named_dev <- setNames(correlations_dev, 
                                   c("Adult_Mortality", "infant_deaths", "Alcohol", 
                                     "percentage_expenditure", "Hepatitis_B", "Measles", 
                                     "BMI", "under_five_deaths", "Polio", "Total_expenditure", 
                                     "Diphtheria", "HIV/AIDS", "GDP", "Population", 
                                     "thinness_1_19_years", "thinness_5_9_years", 
                                     "Income_composition_of_resources", "Schooling"))

correlations_named_devp <- setNames(correlations_devp, 
                                    c("Adult_Mortality", "infant_deaths", "Alcohol", 
                                      "percentage_expenditure", "Hepatitis_B", "Measles", 
                                      "BMI", "under_five_deaths", "Polio", "Total_expenditure", 
                                      "Diphtheria", "HIV/AIDS", "GDP", "Population", 
                                      "thinness_1_19_years", "thinness_5_9_years", 
                                      "Income_composition_of_resources", "Schooling"))

correlations_tbl_dev <- enframe(correlations_named_dev, name = "Variable", value = "Correlation_Developed") %>%
  arrange(desc(Correlation_Developed))

correlations_tbl_devp <- enframe(correlations_named_devp, name = "Variable", value = "Correlation_Developing") %>%
  arrange(desc(Correlation_Developing))
correlations_tbl_dev
correlations_tbl_devp


correlation_comparison <- left_join(correlations_tbl_dev, correlations_tbl_devp, by = "Variable")



cat("\n--- Comparison of Correlations with Life Expectancy: Developed vs Developing Countries ---\n")
print(correlation_comparison)

```

For developed countries, income, percentage expenditure, and GDP showed positive, strong correlation with life expectancy. Polio, BMI, diphtheria and under-five-deaths all had a weak correlation with life expectancy with Pearson coefficients \~0.

in contrast, developing countries had a strong, positive correlation between life expectancy and schooling, income composition index, BMI, GDP, and percentage expenditure, showing that for developing countries, BMI is still an important factor that plays a role in life expectancy, but that this correlation strength decreases as a country becomes a developed country. Percentage of government expenditure on health had a weak correlation with life expectancy in developing countries, likely due to the fact that other factors (poverty, malnutrition, sanitation, etc.) may have a stronger impact on life expectancy, healthcare expenditure may not be evenly distributed across the country, and current government focus is on addressing infectious diseases/emergency healthcare.

Interestingly, total-expenditure had a week correlation in developing countries, and various diseases and alcohol had a mildly strong correlation of around 0.2. Increased years in schooling were strongly associated with higher life expectancy in developing countries, but only moderately so in developed countries.

#### Linear Regression Models by country status

```{r}
#developed
model_developed <- lm(Life_expectancy ~ GDP + Schooling + BMI + Alcohol + Adult_Mortality + 
                        infant_deaths + Polio + Total_expenditure + Diphtheria + Income_composition_of_resources, 
                      data = developed)

#developing
model_developing <- lm(Life_expectancy ~ GDP + Schooling + BMI + Alcohol + Adult_Mortality + 
                         infant_deaths + Polio + Total_expenditure + Diphtheria + 
                         Income_composition_of_resources, 
                       data = developing)

summary(model_developed)


summary(model_developing)
```

#### Testing Interaction Effects of Interest

```{r}
model_dev_int <- lm(Life_expectancy ~ GDP + Schooling + BMI + Alcohol + Adult_Mortality + 
                        infant_deaths + Polio + Total_expenditure + Diphtheria + Income_composition_of_resources + GDP*Schooling, 
                      data = developed)
model_dvl_int <- lm(Life_expectancy ~ GDP + Schooling + BMI + Alcohol + Adult_Mortality + 
                        infant_deaths + Polio + Total_expenditure + Diphtheria + Income_composition_of_resources + GDP*Schooling, 
                      data = developing)
anova(model_developed, model_dev_int)
anova(model_developing, model_dvl_int)

```

Interaction effect of GDP and Schooling does not significantly improve model fit in either model; pval \> 0.05.

```{r}
model_dev_int2 <- lm(Life_expectancy ~ GDP + Schooling + BMI + Alcohol + Adult_Mortality + 
                        infant_deaths + Polio + Total_expenditure + Diphtheria + Income_composition_of_resources + Income_composition_of_resources*Schooling, 
                      data = developed)
model_dvl_int2 <- lm(Life_expectancy ~ GDP + Schooling + BMI + Alcohol + Adult_Mortality + 
                        infant_deaths + Polio + Total_expenditure + Diphtheria + Income_composition_of_resources + Income_composition_of_resources*Schooling, 
                      data = developing)
anova(model_developed, model_dev_int2)
anova(model_developing, model_dvl_int2)

```

Interaction effect of Income Composition Index and Schooling does not significantly improve model fit in either model; pval \> 0.05.

Interested in lifestyle:

```{r}
model_dev_int3 <- lm(Life_expectancy ~ GDP + Schooling + BMI + Alcohol + Adult_Mortality + 
                        infant_deaths + Polio + Total_expenditure + Diphtheria + Income_composition_of_resources + BMI*Alcohol, 
                      data = developed)
model_dvl_int3 <- lm(Life_expectancy ~ GDP + Schooling + BMI + Alcohol + Adult_Mortality + 
                        infant_deaths + Polio + Total_expenditure + Diphtheria + Income_composition_of_resources + BMI*Alcohol, 
                      data = developing)
anova(model_developed, model_dev_int3)
anova(model_developing, model_dvl_int3)
summary(model_developing)$r.squared
summary(model_dvl_int3)$r.squared
```

Interaction effect of BMI and Alcohol does not significantly improve model fit in developed countries, but it is significant in developing countries with a p-value of 0.02, less than the threshold of 0.05.

-   This suggests that the relationship between **alcohol consumption** and **BMI** in relation to **life expectancy** might differ between developed and developing countries.

-   R\^2 increased minutely, indicating better model fit

-   Countries with lower avg BMI experience less negative impact of alcohol consumption on life expectancy, whereas those with higher BMI may experience a greater reduction in life expectancy at similar levels of alcohol consumption.

**Further exploration of interaction effect:**

```{r}
developing |>
  ggplot(aes(x = Alcohol, y = log(BMI))) +
  geom_point(alpha = 0.7) + 
  geom_smooth()
```

```{r}
developing$predicted_life_expectancy <- predict(model_dvl_int3, newdata = developing)

ggplot(developing, aes(x = Alcohol, y = predicted_life_expectancy, color = BMI)) +
  geom_point() + 
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Interaction Effect: Alcohol and BMI on Life Expectancy",
       x = "Alcohol Consumption", y = "Predicted Life Expectancy")

```

```{r}
model_dev_int4 <- lm(Life_expectancy ~ GDP + Schooling + BMI + Alcohol + Adult_Mortality + 
                        infant_deaths + Polio + Total_expenditure + Diphtheria + Income_composition_of_resources + Total_expenditure*Polio, 
                      data = developed)
model_dvl_int4 <- lm(Life_expectancy ~ GDP + Schooling + BMI + Alcohol + Adult_Mortality + 
                        infant_deaths + Polio + Total_expenditure + Diphtheria + Income_composition_of_resources + Total_expenditure*Polio, 
                      data = developing)
anova(model_developed, model_dev_int4)
anova(model_developing, model_dvl_int4)
```

Interaction effect of total expenditure and Polio does not significantly improve model fit in either model; pval \> 0.05.

#### Assessing Final Model (including interaction effect BMI\*Alcohol)

```{r}
# install.packages("car")
library(car)
vif(model_dvl_int3, type = "predictor")
model_dvl_int3
```

None have GVIF \> 10, so multicollinearity is not an issue with this model

#### Checking Model Assumptions

Constant Variance assumption satisfied.

```{r}

residuals_dev <- residuals(model_dev_int3)
fitted_values_dev <- fitted(model_dev_int3)


plot(fitted_values_dev, residuals_dev, 
     xlab = "Fitted Values", 
     ylab = "Residuals", 
     main = "Residuals vs Fitted Values for Developed Model")
abline(h = 0, col = "red")

residuals_dvl <- residuals(model_dvl_int3)
fitted_values_dvl <- fitted(model_dvl_int3)


plot(fitted_values_dvl, residuals_dvl, 
     xlab = "Fitted Values", 
     ylab = "Residuals", 
     main = "Residuals vs Fitted Values for Developing Model")
abline(h = 0, col = "red")
```

Normality Assumption satisfied; residuals are normally distributed.

```{r}

hist(residuals_dev, 
     main = "Histogram of Residuals for Developed Countries", 
     xlab = "Residuals", 
     breaks = 20)
hist(residuals_dvl, 
     main = "Histogram of Residuals for Developing Countries", 
     xlab = "Residuals", 
     breaks = 20)
```

## Conclusion

Emphasis should be placed on:

1.  **Education Programs:** Better education results in improved public health awareness, lifestyle choices, and access to resources.

-   Investments in education systems with a rural/underserved focus

-   Teacher training programs

-   Collaboration w/ national education ministries

-   Health expenditure monitoring program to fairly distribute resources

2.  **Economic Development:** Investments aimed at economic growth and resource allocation would significantly positively impact life expectancy.

-   Develop Global Income Composition Index and GDP Goals that countries have incentive to meet

-   Promote health-sensitive economic policies, health financing models

-   Support mobilization of resources through taxation policy, international health funding

3.  **Health System Infrastructure:** Strengthening weak correlations between total health expenditure and life expectancy

-   Make sure health spending directly benefits the population through efficient healthcare delivery

-   Address inequalities in health system and target vulnerable populations

-   Supporting disease prevention programs

-   Increasing vaccination support and infectious disease research for diphtheria, polio, etc.

## Limitations

Independence of observations is not met, since there could be interdependence between countries due to geographic proximity, trade relations, shared economic conditions, or regional policies. However, I proceeded with the analysis because:

-   Model was built with control for observable characteristics, or relevant covariates that account for country-specific differences (GDP, population, etc)., reducing risk of bias due to country-specific interdependencies

-   Time period of interest is relatively short, spanning from 2000 to 2015, so temporal autocorrelation (correlation across years within a country) is limited

Additional limitations include omission of NAs from dataset, which may lead to differences in results.

## Future Work

-   **Inclusion of fixed effects** (year-level) in linear regression to control for time-invariant characteristics that could affect the relationship between the variables

```{=html}
<!-- -->
```
-   Comparison between results with NAs and without NAs

-   **Time-Series Analysis** w/ ARIMA to determine if there are any trends over time
